Discovery of a Heparan-Sulfate Binding Domain in Monkeypox Virus H3 Protein as a New Anti-poxviral Drug Target

Version 1

Abstract

Abstract

The 2022 global outbreak of mpox virus (MPXV), recognized as a public health emergency by the World Health Organization (WHO), underscored the urgent need for specific antiviral agents. H3, a key antiviral target, mediates the adhesion of MPXV to host cells through its interaction with heparan sulfate (HS). Understanding the details of the H3-HS interaction has been challenging due to the highly flexible and complex nature of HS, which hinders its structural characterization. Utilizing state-of-the-art computational biology techniques, including molecular dynamics (MD) simulations, dynamic molecular docking, and AlphaFold2, we identified a novel α-helical domain in H3 critical for HS binding. This domain, conserved across orthopoxviruses, was validated experimentally, confirming its crucial role in viral adhesion at both molecular and cellular levels. Targeting this newly discovered HS-binding domain, we de novo designed a protein inhibitor, AI-PoxBlock723, which effectively disrupt the H3-HS interaction, and inhibit viral infection. Our work not only introduces a promising anti-poxviral drug candidate, but also provides a comprehensive framework for elucidating complex protein-glycan interactions.

Introduction

Mpox virus (also known as monkeypox virus), a member of the Orthopoxvirus genus within the Poxviridae family, is an emerging zoonotic pathogen. This genus also includes the Variola virus (VARV), the causative agent of smallpox, and Vaccinia virus (VACV), which has been used as a live vaccine against smallpox1,2. Historically endemic to Central and West Africa, MPXV gained global attention in 2022 when it spread internationally, prompting the World Health Organization (WHO) to declare it a public health emergency of international concern3,4. Unlike smallpox, MPXV has a lower fatality rate of 2-10%, yet it continues to pose significant health challenges due to the absence of specific antiviral treatments5–10.

The adhesion of MPXV to host cells, mediated through the viral adhesion protein, represents the critical first step in viral infection11–13. For example, H3 binds to cell-surface heparan sulfate (HS), which is not only pivotal for MPXV but also represents a potential universal target across orthopoxviruses. Therefore, H3-specific neutralizing monoclonal antibody protects rabbits and immunodeficient mice against lethal poxvirus infections14, and novel mpox vaccines have been developed recently with H3 as an important component15,16. However, the structural intricacies of the H3-HS interaction are not well-understood, largely due to the dynamic and heterogeneous nature of HS. This has significantly hindered the development of targeted therapeutics.

Recent advances in computational biology have opened new avenues for elucidating complex biological interactions. This study employs state-of-the-art AI-enhanced structural predictions, all-atom molecular dynamics simulations, and dynamic molecular docking to uncover novel insights into the H3-HS interaction. We identified here a previously uncharacterized α-helical domain within H3 which is essential for effective HS binding17–21. Further biophysical and biochemical validation through single-molecule force spectroscopy, flow cytometry and biolayer interferometry has confirmed the functional importance of this domain in the viral adhesion process22–25. Based on this discovery, we developed a de novo designed protein inhibitor targeting the α-helical domain of H3, demonstrating efficacy in viral inhibition assays26,27. These findings not only highlight a promising therapeutic approach against MPXV and potentially other pathogens utilizing similar adhesion mechanisms but also demonstrate the powerful capabilities of computational biology, particularly AI-assisted methodologies, in the discovery of specific antiviral agents..

Result

Identification of the Novel HS Binding Site through Comprehensive Computational Biology Methods

The H3 protein of MPXV is a 324-amino-acid transmembrane protein and conserved across all orthopoxviruses. Pioneer experiments have shown that its extracellular domain (residues 1-284) binds to cell-surface heparan sulfate (HS)28,29, facilitating the viral particle’s entry in conjunction with the fusion protein system30. Therefore, inhibiting HS binding to H3 represents a promising strategy for anti-poxviral drug development. Although the structure of the extracellular region of VACV H3 is known, the dynamic and heterogeneous nature of HS complicates the identification of precise binding sites and mechanisms, challenging the development of targeted antiviral treatments.

To elucidate potential HS binding sites, we initially analyzed the sequence and structure of H3. The AI-based AlphaFold2 has proven to be a powerful tool for protein structure prediction and is increasingly used for protein-target drug development. Given that only the H3 structure from VACV was available28, we utilized AlphaFold2 to model the MPXV H3 structure (Fig. S1), focusing on regions rich in basic amino acids—K (Lysine), R (Arginine), and H (Histidine)—which are known to interact with glycosaminoglycans (GAG) like HS. We identified three potential HS binding motifs first: Motif 1 (K95, R96, R100), Motif 2 (K141, K146, K147), and Motif 3 (K161, K163)31, each located within the known homologous structure (VACV H3) (Fig. 1C, D).

Structural Analysis of the H3 Protein in Monkeypox Virus

(A) Phylogenetic tree depicts the evolutionary relationships within the Poxviridae family, highlighting MPXV, (blue circle), VARV, (red circle), and VACV, (green circle). (B) Schematic shows MPXV adhesion to a cell, with H3 and fusion complexes depicted. (C) The amino acid sequence and structure of MPXV H3 displays β-strands (yellow arrows) and α-helices (pink cylinders). Mg(II) binding sites (green) and potential HS binding motifs (blue underlines) are shown. The novel three-helical domain (240-282), is highlighted in yellow. (D) AlphaFold2 prediction of MPXV H3 Structure shownin two orientations. Regions matching VACV H3’s known structure are in blue; novel elements predicted are in yellow, with all potential GAG-binding motifs highlighted.

Interestingly, a cluster of seven additional positively charged residues (R242, H247, R248, K253, R259, K266, R267) in the C-terminal region was revealed. This region, not structurally characterized previously due to its removal in X-ray diffraction studies, was retained in all other biochemical and viral assays28–30. Alphafold2 predicted these residues to form a three-helical domain, which may represent a novel binding site for HS (Fig. 1C, D).

To validate these predicted binding sites, we employed flexible molecular docking followed by extensive molecular dynamics (MD) simulations32,33. We selected a 20-repeat HS unit with a specific composition for detailed analysis ([IdoA2S-GlcNS6S-IdoA-GlcNS(3,6S)]5) (Fig. S2). A 500 ns MD simulation established the equilibrium conformation of H3, and subsequent docking with AutoDock Vina identified four high-scoring conformations for each motif (Fig. 2A, Fig. S3). To address the dynamic nature of HS, we extended our simulations to 3 times additional 1000 ns MD simulations for each conformation. RMSD (Root Mean Square Deviation) analysis demonstrated notably stable binding for HS to both Motif 1 and the newly identified helical domain (Supplementary Video 1), with RMSD values consistently below 4 nm, indicating minimal structural fluctuations during interaction (Fig. 2B).

Molecular Dynamics and Docking Analyses of H3-HS Interactions

(A-B) Cartoons show docking results of heparan sulfate (HS) with H3 motifs 1, 2, 3, and the helical domain, respectively. Panels (B) show RMSD values from 1µs MD simulations, color-coded to match the configurations in (A). (C) Schematics illustrates the reaction coordinate in umbrella sampling, highlighting HS dissociation from H3 with a green force curve. (D) Presents binding free energies for HS-H3 interactions in motifs 1, 2, 3, and the helical region, color-coded respectively. (E) A free energy landscape map from a 1000 ns REMD simulation shows HS binding configurations to the helical and Mg(Ⅱ) regions. (F) Panel provides salt bridge formation statistics between HS and H3’s basic amino acids, with a bar chart of average formations. (G) A surface plot shows the frequency of salt bridge formations within H3, with areas of frequent formations in blue. (H) Detailed views of HS-H3 interactions, with the left image showing salt bridges and the right image displaying electrostatic interactions with Mg(Ⅱ). This panel illustrates the impact on HS binding stability to H3 following the removal of Mg(Ⅱ) during the simulation. (J) The effects of mutating all basic amino acids in the helical domain on the binding stability of HS are shown. (K) The “palm-binding” model is depicted where HS is secured by the helical “fingers” and interacts with the Mg(Ⅱ) “palm.”

Moreover, we quantitatively analyzed the binding free energy between H3 and HS for different conformational states using umbrella sampling techniques. A harmonic potential was applied to HS and gradually pulled away from H3, generating a series of reaction coordinates (Fig. 2C, Fig. S4, Supplementary Video 2). The green force-extension curve represents the force during this stretching process. This curve, along with the corresponding PMF (Potential of Mean Force) variation shown in Figure S4, demonstrated the free energy changes throughout the umbrella sampling process, revealing that the helical domain possesses the highest binding affinity with a ΔG of -45 kCal/mol (Fig. 2D).

Although the exact binding site between H3-HS is unknown, previous studies have suggested the critical role of the Mg(II) ion in stabilizing HS binding, where its removal reduces binding efficiency28. Indeed, our newly discovered α-helical domain is very close to the Mg(Ⅱ)-bound region in H3. Notably, these two parts appear to form a distinct cavity-like binding pocket for HS (Fig. 1D, Fig. 2E). Thus, we conducted a 1000 ns REMD (Replica Exchange Molecular Dynamics) simulation of the H3-HS complex to further understand the binding dynamics. REMD, an advanced sampling technique, helps overcome the kinetic barriers typically encountered in conformational transitions by utilizing a series of temperature-controlled replicas to enhance the exploration of the conformational landscape34. For this study, we set up 64 replicas over a temperature range from 310K to 387K, cumulatively simulating for 64 µs. This comprehensive simulation captured the HS molecule progressively binding deeper into a cavity formed by the α-helical domain, eventually interacting with the Mg(II) ion as hypothesized (Fig. 2E, Supplementary Video 3).

The free energy landscape, mapped out from these simulations, shows HS transitioning from a stable interaction with the helical domain across an energy barrier into the Mg(II)-enhanced cavity. This dual interaction—first with the helical domain and then with the Mg(II) site—was consistently observed to stabilize the complex further, evidenced by a significantly lower binding free energy of -67 kCal/mol (Fig. S4)35. These results underscore the dynamic nature of the H3-HS interaction and validate our model of sequential binding, which could be critical for designing inhibitors that target these specific interactions.

Given the crucial role of electrostatic interactions in the HS-H3 binding process, we conducted an exhaustive analysis of salt bridge formations across all MD simulation trajectories. This analysis focused on the formation of salt bridges over time between the basic amino acids (R, K, H) of H3 and HS, with brighter areas indicating more frequent formation of salt bridges and robust interaction (Fig. 2F, Fig. S5). A bar graph accompanied these maps, quantifying the average number of salt bridges each amino acid formed during the simulations, further illustrating the intensity and frequency of these interactions. Our findings demonstrated a pronounced affinity of HS for the helical domain of H3, with basic amino acids in this region forming significantly more salt bridges compared to other parts of the protein. The surface hotspot map of H3 (Fig. 2G) visually highlighted these interactions, with regions of frequent salt bridge formation represented in varying shades of blue, underscoring their distribution and intensity. In the helical domain, a focus view showed that the 1st, 7th, 9th, and 18th sugar units of HS formed salt bridges with R248, R242, R267, and R259 of H3, respectively (Fig. 2H, left), while the sulfate group on the 2nd sugar side chain of HS bound to Mg(II) (Fig. 2H, right).

To further substantiate these findings, we performed two control experiments. First, we removed the Mg(Ⅱ) for MD simulation. Based on the results from umbrella sampling, we extended the simulation of the most stable H3-HS binding conformation to 1000 ns, then removed the Mg(Ⅱ) ion and continued the simulations. The resulting increase in fluctuations of the RMSD values confirmed the stabilizing role of Mg(Ⅱ) in the binding process (Fig. 2I, Supplementary Video 4). Second, we mutated all seven basic amino acids in the helical domain to serine and observed significant alterations in HS binding dynamics, supporting the critical role of these positively charged residues in stabilizing the interaction (Fig. 2J, Supplementary Video 5).

Thus, the interaction model revealed that the α-helical domain of H3 functions like a thumb, guiding HS into a stable binding position, while the Mg(Ⅱ) region acts like a palm, securing the HS in place. This dynamic interaction allows HS to transition from initially binding at the protein surface to being deeply anchored within the protein cavity, illustrating a complex but well-organized binding process (Fig. 2K).

Analysis of the H3 protein sequence across the Poxviridae family highlights its conservation and significance in HS binding. Using NCBI’s BLAST, we retrieved sequences of H3 from 66 Poxviridae members and performed multiple sequence comparisons36. Structural predictions via AlphaFold2 demonstrated that these sequences generally share a similar architecture to MPXV H3, particularly the presence of the α-helical domain (Fig. S6). Given the crucial role of electrostatic interactions in the H3-HS binding process, we further analyzed the charge distribution within this domain. A heatmap detailing the side-chain charge distribution at a physiological pH of 6.5—mimicking the microenvironment of H3-HS interaction— revealed a significant concentration of positive charges at the C-terminal end of the helical domain (Fig. 3A).

Charge Characteristics and Structural Analysis of H3 Protein

(A) The heatmap shows the amino acid charge distribution in the H3 protein across 66 Poxviridae viruses, following multiple sequence alignment. Blue indicates areas with more positive charges, and red indicates more negative charges. The accompanying curve shows the average charge of all amino acids in Poxviridae H3. (B) Logo plot of the amino acid sequence of the helical region, highlighting the conservation of basic amino acids at specific positions. (C) The surface charge analysis of H3 from the MPXV, VACV, VARV, and cowpox viruses (CPXV) with the helical domain showing a significantly positive charge. (D) Schematic of the single-molecule force spectroscopy unfolding experiment on H3, illustrating the unfolding process of the helical region (yellow) followed by the main body (blue). (E) Representative curves of H3 unfolding, color-coded to show the helical region (yellow), main body (blue), and full-length (purple) unfolding. (F) Histograms depict the force spectroscopy signals for helical domain, main body, and full-length unfolding, with ΔLc statistics provided. The inset shows a Gaussian fit of unfolding forces.

Notably, this domain consistently exhibited an accumulation of basic amino acids such as lysine, arginine, and histidine, which are essential for binding the negatively charged HS. These amino acids, including residues identified as 358, 361, 367, 373, 379, 386, and 387 in Poxviridae H3 (global position in alignment), corresponding to residues H239, R242, R248, K253, R259, K266, R267 in MPXV H3, show high conservation (Fig. 3B, S7). Surface charge analysis of H3 proteins from four orthopoxviruses, including MPXV, VACV, VARV and CPXV confirmed the predominance of strong positive charges in their α-helical domains (Fig. 3C), crucial for effective binding to HS. This consistent feature across different viruses underscores the evolutionary importance of the helical domain, suggesting a universal mechanism in orthopoxviral adhesion that could be targeted in antiviral strategy.

Experimental Confirmation of the Structure and Function of α-Helical Domain

To validate our computational findings, we first conducted experiments using atomic force microscopy-based single-molecule force spectroscopy (AFM-SMFS). This technique, widely used for studying protein (un)folding, allowed us to directly assess the structure and stability of the α-helical domain within the H3 protein37–39. We engineered the H3 construct with two GB1 domains, each 18 nm in length upon unfolding, and one Cohesin module for precise measurement (Fig. S8)40–42. During the experiment, the coated AFM tip with GB1 and Dockerin initiated a Coh-Doc interaction, producing a characteristic sawtooth-like force-extension curve as it was stretched (Fig. 3D, 3E). Notably, two peaks corresponding to the unfolding of the α-helical domain were observed. The first peak showed a contour length increment (ΔLc) of 13 nm, closely matching the theoretical unfolding length for the 42 amino acid-long α-helical domain of H3 (residue 282-240, 42*0.36 nm/aa-2.6 nm)43. The measured unfolding force was 67.4 ± 2.1 pN (n=139), while the second peak indicated the unfolding of the remaining structure of H3, with a ΔLc of 84 nm (Fig. 3F). The total unfolding, represented by a cumulative ΔLc of 97 nm, confirmed the structured and stable nature of the α-helical domain within H3.

We further explored the functional role of the α-helical domain in mediating H3’s binding to HS using AFM on live cells44,45. The H3 protein was linked to the AFM tip and approached cultured CHO-K1 cells, which express HS, mounted on a Petri dish. Using inverted microscopy for precise positioning, the interaction between H3 and HS was initiated upon contact, subsequently monitored by retracting the tip to generate a force-extension curve. This process highlighted a distinct peak indicative of H3-HS binding with a dissociation force of 33.7 ± 0.2 pN and a binding probability of 43% (Fig. 4A, B). Trials performed across different cell surface areas produced a detailed force map, quantifying the distribution of HS unbinding forces and confirming the essential role of the helical domain in viral adhesion (Fig. 4C, D).

Analysis of the Helical Domain’s Interaction with HS at Cellular Level

(A) Schematic of the cell force spectroscopy experiment setup shows three scenarios: wild-type H3 on an AFM tip interacting with HS on CHO-K1 cells, mutation of all basic amino acids in the H3 helical region to serine, and cells treated with HS hydrolase to remove surface HS before testing. (B) Force-extension curves depict interactions for the wild type, mutant, and control groups, marked with blue and red asterisks for dissociation events. An inset shows the optical microscope positioning the AFM probe. (C) Histograms of dissociation signals comparing the wild type, mutant, and control groups, with an inset detailing the surface distribution of dissociation forces. (D) Statistical graph showing binding probabilities for different groups, highlighting significant differences determined by t-tests (p<1e-5). (E) Flow cytometry results illustrate interactions of wild-type (WT) and Uncharged H3 fused with eGFP with CHO-K1 cells, alongside surface HS removed control (green) and cell-only control (Blank, grey). (F) Statistical analysis of flow cytometry data, showing significant differences between groups as determined by t-tests. *****, P<1e-5

To explore the α-helical domain’s role in binding HS, we engineered a variant of the H3 protein, named H3(uncharged), in which all positively charged residues were replaced with the uncharged amino acid serine. AFM experiments on this variant showed a detectable force peak with a lower unbinding force of 28.8 ± 0.2 pN, compared to 33.7 ± 0.3 pN for the wild type, along with a decreased HS binding probability from 43% to 25% (Fig. 4C, D). These findings indicate that the basic amino acids in the helical domain are crucial for HS binding. Additionally, to ensure the specificity of the interaction signals, we treated CHO-K1 cells with heparinase II, which degrades HS. Control experiments using this enzyme treatment demonstrated a further reduction in the dissociation force to 23.0 pN and a decrease in binding probability to 12.6%, confirming that the observed interactions specifically originated from the dissociation of the HS-H3 complex.

Further validation was carried out using flow cytometry (FCM). We fused eGFP, a fluorescent reporter protein, to the C-terminus of both the wild-type H3 (H3(WT)) and H3(uncharged). Following incubation of fusion proteins with cells, FCM analysis demonstrated a significant reduction in mean fluorescence intensity in cells treated with H3(uncharged) compared to H3(WT) (Fig. 4E). Statistical analysis of multiple experiments highlighted a significant difference between the two variants (Fig. 4F). These results reinforce the findings from the MD simulations and force spectroscopy, emphasizing that the positively charged amino acids within the helical domain of H3 are critical for the efficient binding to HS.

Design Protein Inhibitor for H3 Adhesion Targeting the Helical Domain

Following the discovery that the helical domain is crucial for the binding of H3 to HS, we focused on designing specific inhibitors to target this domain, aiming to both validate its functional significance and explore potential antiviral therapies using recently developed powerful AI tools for protein. Utilizing RFdiffusion, we designed a series of inhibitors with α-helical structures and named it AI-PoxBlock (Fig. 5A). Sequence recovery was performed on 1000 backbones using ProteinMPNN, generating ten sequences per backbone. These designs were further validated through complex structure predictions using AlphaFold2, selecting candidates based on pAE values below 7 and iPTM scores above 0.9. Additionally, the predicted structures of these candidates exhibited RMSD values less than 2 nm when compared to the RFdiffusion-generated structures (Fig. S9). The interactions between the helical domain of H3 and these inhibitors were then examined in 500 ns MD simulations, confirming binding capabilities through the analysis of RMSD trajectories, where five inhibitors—AI-PoxBlock302, 602, 614, 723, and 761—showed stable RMSD values under 0.5 nm (Fig. S10, S11, Supplementary Video 6).

Development and Testing of Protein Inhibitors Targeting the H3 Helical Domain

(A) Diagrams depict protein inhibitors designed to target the H3 helical region, created using RFdiffusion. Sequences capable of folding into the target scaffold structures were generated using ProteinMPNN, and were validated through AlphaFold2, followed by 500 ns MD simulations for structural stability and interaction scoring. (B) FCM analysis demonstrates the inhibitory effect of AI-Poxblock723 at various concentrations. (C) BLI confirms direct interaction between AI-Poxblock723 and the H3 helical domain. (D) Graphs display the inhibitory effect of the indicated AI-Poxblocks on MPXV infection of Vero E6 cells, with quantitative analysis of virus-infected foci provided on the right.

Although AI-PoxBlock302 could not be successfully expressed, the remaining four inhibitors underwent further testing in FCM to evaluate their efficacy in inhibiting H3 binding to CHO-K1 cell surfaces. AI-PoxBlock723 stood out, significantly reducing H3 binding at a concentration of 10 μM, while the other inhibitors displayed no tendency to inhibit H3 binding to HS on cell surfaces (Fig. 5B, Fig. S12). Further analysis using Biolayer Interferometry (BLI) confirmed that AI-PoxBlock723 binds to H3 with an equilibrium dissociation constant (KD) of 9 μM (Fig. 5C). To ensure specificity, we also assessed the interaction between AI-PoxBlock723 and a truncated version of H3 (residues 1-239), lacking the α-helical domain. BLI results indicated a marked reduction in binding, reinforcing the specificity of AI-PoxBlock723 for the helical domain (Fig. S13). Circular Dichroism (CD) spectroscopy confirmed that AI-PoxBlock723 predominantly consists of α-helices, aligning with its design specifications (Fig. S14). To evaluate the antiviral activity of AI-PoxBlocks, they were series diluted and incubated with MPXV before the infection of Vero E6 cells and cultured overnight, followed by the determination of virus infection foci as we previously described46. Excitingly, AI-PoxBlock723 showed a significant antiviral effect with the half maximal inhibitory concentration (IC50) of 8.86 μM, whereas other AI-PoxBlocks exhibited no inhibitory efficiency. (Fig. 5D). Considering the high conservation of H3, particularly its helical domain among the four orthopoxviruses pathogenic to humans (Fig. S15), we also assessed the activity of AI-PoxBlocks against VACV infection of target cells. Consistently, AI-PoxBlock723 is also effective for inhibiting VACA infection in vitro with a similar IC50 of 8.67μM (Fig. S16), suggesting the helical domain as an antiviral drug pocket is universal at least for orthopoxvirus. Collectively, these results demonstrate the efficacy of the designed inhibitors targeting the H3 helical domain and validate the crucial role of this domain in viral adhesion, providing a promising foundation for the development of novel antiviral agents.

Discussion

Our study provides detailed mechanistic insights into the adhesion process of the MPXV, emphasizing the role of the H3 protein as a crucial adhesive component initiating host-cell interactions in the virus’s lifecycle. By integrating computational methods with experimental validations, we have identified a previously unknown α-helical domain within the H3 protein that is essential for binding to HS on the host cell surface. This discovery enhances our understanding of MPXV pathogenesis and presents a novel target for therapeutic intervention.

HS serves as a cellular ligand for various viruses47–49, underscoring the importance of studying its interactions with viral adhesion proteins. However, the complexity and flexibility of polysaccharides like HS have historically posed challenges for the structural analysis of protein-polysaccharide complexes48,50,51. This complexity has rendered the molecular details of the interaction between MPXV’s H3 and host cell HS elusive. By employing MD simulations and AI-based structural prediction tools such as AlphaFold2, we have overcome these obstacles, revealing key interaction sites—specifically the α-helical domain—and binding mechanisms16,52–59. Our study demonstrates the feasibility of our comprehensive computational approach in researching viral proteins involved in protein-glycan interactions, offering a promising methodological advancement.

Looking forward, the methodologies and findings from this study are poised to catalyze further investigations into orthopoxviruses and other viral families employing similar adhesion strategies. The development of innovative protein inhibitors, driven by state-of-the-art AI models for protein design including RFdiffusion, ProteinMPNN, and AlphaFold2, highlights the transformative impact of AI in drug discovery. Future research could focus on refining the efficacy and specificity of inhibitors targeting the α-helical domain and explore their clinical applicability60. Extending our integrative approach to other viral systems could reveal additional therapeutic targets, significantly enhancing our capacity to combat viral diseases.

In summary, our research provides a comprehensive computational methodology to study complex protein-glycan interactions and paves the way for the development of innovative antiviral drugs. These findings offer promising potential treatments for MPXV and other severe viral diseases, demonstrating the powerful capabilities of AI-assisted computational biology in the discovery of specific antiviral agents.

eLife Assessment

This work presents important findings regarding the interaction of the monkeypox virus (MPXV) attachment H3 protein with the cellular receptor heparan sulfate and the use of this information to develop antivirals potentially effective against all orthopoxviruses. Using a combination of state-of-the art computational and wet experiments the authors present solid evidence to sustain their claims. These results will interest those working on basic orthopoxviruses biology and antiviral development.

Reviewer #1 (Public Review):

Summary:

The study aimed to better understand the role of the H3 protein of the Monkeypox virus (MPXV) in host cell adhesion, identifying a crucial α-helical domain for interaction with heparan sulfate (HS). Using a combination of advanced computational simulations and experimental validations, the authors discovered that this domain is essential for viral adhesion and potentially a new target for developing antiviral therapies.

Strengths:

The study's main strengths include the use of cutting-edge computational tools such as AlphaFold2 and molecular dynamics simulations, combined with robust experimental techniques like single-molecule force spectroscopy and flow cytometry. These methods provided a detailed and reliable view of the interactions between the H3 protein and HS. The study also highlighted the importance of the α-helical domain's electric charge and the influence of the Mg(II) ion in stabilizing this interaction. The work's impact on the field is significant, offering new perspectives for developing antiviral treatments for MPXV and potentially other viruses with similar adhesion mechanisms. The provided methods and data are highly useful for researchers working with viral proteins and protein-polysaccharide interactions, offering a solid foundation for future investigations and therapeutic innovations.

Weaknesses:

However, some limitations are notable. Despite the robust use of computational methodologies, the limitations of this approach are not discussed, such as potential sources of error, standard deviation rates, and known controls for the H3 protein to justify the claims. Additionally, validations with methodologies like X-ray crystallography would further benefit the visualization of the H3 and HS interaction.

Reviewer #2 (Public Review):

Summary:

The manuscript presenting the discovery of a heparan-sulfate (HS) binding domain in monkeypox virus (MPXV) H3 protein as a new anti-poxviral drug target, presented by Bin Zhen and co-workers, is of interest, given that it offers a potentially broad antiviral substance to be used against poxviruses. Using new computational biology techniques, the authors identified a new alpha-helical domain in the H3 protein, which interacts with cell surface HS, and this domain seems to be crucial for H3-HS interaction. Given that this domain is conserved across orthopoxviruses, authors designed protein inhibitors. One of these inhibitors, AI-PoxBlock723, effectively disrupted the H3-HS interaction and inhibited infection with Monkeypox virus and Vaccinia virus. The presented data should be of interest to a diverse audience, given the possibility of an effective anti-poxviral drug.

Strengths:

In my opinion, the experiments done in this work were well-planned and executed. The authors put together several computational methods, to design poxvirus inhibitor molecules, and then they test these molecules for infection inhibition.

Weaknesses:

One thing that could be improved, is the presentation of results, to make them more easily understandable to readers, who may not be experts in protein modeling programs. For example, figures should be self-explanatory and understood on their own, without the need to revise text. Therefore, the figure legend should be more informative as to how the experiments were done.

Reviewer #3 (Public Review):

Summary:

The article is an interesting approach to determining the MPOX receptor using "in silico" tools. The results show the presence of two regions of the H3 protein with a high probability of being involved in the interaction with the HS cell receptor. However, the α-helical region seems to be the most probable, since modifications in this region affect the virus binding to the HS receptor.

Strengths:

In my opinion, it is an informative article with interesting results, generated by a combination of "in silico" and wet science to test the theoretical results. This is a strong point of the article.

Weaknesses:

Has a crystal structure of the H3 protein been reported?

The following text is in line 104: "which may represent a novel binding site for HS". It is unclear whether this means this "new binding site" is an alternative site to an old one or whether it is the true binding site that had not been previously elucidated.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

The study aimed to better understand the role of the H3 protein of the Monkeypox virus (MPXV) in host cell adhesion, identifying a crucial α-helical domain for interaction with heparan sulfate (HS). Using a combination of advanced computational simulations and experimental validations, the authors discovered that this domain is essential for viral adhesion and potentially a new target for developing antiviral therapies.

Strengths:

The study's main strengths include the use of cutting-edge computational tools such as AlphaFold2 and molecular dynamics simulations, combined with robust experimental techniques like single-molecule force spectroscopy and flow cytometry. These methods provided a detailed and reliable view of the interactions between the H3 protein and HS. The study also highlighted the importance of the α-helical domain's electric charge and the influence of the Mg(II) ion in stabilizing this interaction. The work's impact on the field is significant, offering new perspectives for developing antiviral treatments for MPXV and potentially other viruses with similar adhesion mechanisms. The provided methods and data are highly useful for researchers working with viral proteins and protein-polysaccharide interactions, offering a solid foundation for future investigations and therapeutic innovations.

Weaknesses:

However, some limitations are notable. Despite the robust use of computational methodologies, the limitations of this approach are not discussed, such as potential sources of error, standard deviation rates, and known controls for the H3 protein to justify the claims. Additionally, validations with methodologies like X-ray crystallography would further benefit the visualization of the H3 and HS interaction.

Thank you very much for the evaluation and appreciation of our work. In response to the identified weakness, we have conducted additional analyses to further assess the limitations of the computational methodologies used. Specifically, we predicted the MPXV H3 structure using two other AI-based protein structure prediction models, ESMFold and RoseTTAFold2. Both models also predicted an a-helical structure, which supports our conclusion. However, they yielded lower pLDDT scores (Figure S1A-C in the revised SI), indicating that some error may be present.

We agree with this reviewer, as well as the other reviewers, that X-ray crystallography data for the H3 structure would be highly valuable. Unfortunately, we lack the expertise in structural biology to obtain these results at this stage. To complement this, we performed molecular dynamics (MD) simulations, which suggest that the helical domain is connected to the main domain via a flexible linker. This flexibility may help explain the challenges in obtaining a high-resolution X-ray structure. In fact, to date, the only structural data available for H3 is from the VAVC, which excludes the helical domain (The helical domain part is cleaved for the X-ray studies). We have added this point to the discussion and hope that experts in structural biology will be able to resolve the structure of this domain in the future.

Reviewer #2 (Public Review):

Summary:

The manuscript presenting the discovery of a heparan-sulfate (HS) binding domain in monkeypox virus (MPXV) H3 protein as a new anti-poxviral drug target, presented by Bin Zhen and co-workers, is of interest, given that it offers a potentially broad antiviral substance to be used against poxviruses. Using new computational biology techniques, the authors identified a new alpha-helical domain in the H3 protein, which interacts with cell surface HS, and this domain seems to be crucial for H3-HS interaction. Given that this domain is conserved across orthopoxviruses, authors designed protein inhibitors. One of these inhibitors, AI-PoxBlock723, effectively disrupted the H3-HS interaction and inhibited infection with Monkeypox virus and Vaccinia virus. The presented data should be of interest to a diverse audience, given the possibility of an effective anti-poxviral drug.

Strengths:

In my opinion, the experiments done in this work were well-planned and executed. The authors put together several computational methods, to design poxvirus inhibitor molecules, and then they test these molecules for infection inhibition.

Weaknesses:

One thing that could be improved, is the presentation of results, to make them more easily understandable to readers, who may not be experts in protein modeling programs. For example, figures should be self-explanatory and understood on their own, without the need to revise text. Therefore, the figure legend should be more informative as to how the experiments were done.

Thank you very much for your appreciation of our work and your support. In response to the identified weakness, we have carefully reviewed all the figure legends to ensure they are more informative.

Reviewer #3 (Public Review):

Summary:

The article is an interesting approach to determining the MPOX receptor using "in silico" tools. The results show the presence of two regions of the H3 protein with a high probability of being involved in the interaction with the HS cell receptor. However, the α-helical region seems to be the most probable, since modifications in this region affect the virus binding to the HS receptor.

Strengths:

In my opinion, it is an informative article with interesting results, generated by a combination of "in silico" and wet science to test the theoretical results. This is a strong point of the article.

Weaknesses:

Has a crystal structure of the H3 protein been reported?

The following text is in line 104: "which may represent a novel binding site for HS". It is unclear whether this means this "new binding site" is an alternative site to an old one or whether it is the true binding site that had not been previously elucidated.

Thank you very much for your thoughtful evaluation and appreciation of our work.

We agree with this reviewer, as well as the other reviewers, that X-ray crystallography data for the H3 structure would be highly valuable. Unfortunately, we are not experts in structural biology, and we have not yet been able to obtain these structural results. To date, the only structure available for H3 is the one from VAVC, which does not include the helical domain. We have included this point in the discussion and hope that experts in structural biology will be able to resolve the structure of this domain in the future.

Regarding the "novel binding site," this term refers to "the true binding site that had not been previously elucidated." Previous research identified that H3 binds to heparan sulfate (HS), but the exact binding site had not been determined.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Validation of Results with Other Experimental Methods: While single-molecule force spectroscopy and flow cytometry provide valuable data, including complementary methods such as X-ray crystallography could offer additional insights into the H3-HS interaction and the effectiveness of the inhibitors.

Discussion of Computational Model Limitations: Although the use of AlphaFold2 and other advanced tools is a strength, it is important to discuss the limitations of these models in more detail, including potential sources of error and how they may impact the interpretation of the results.

During the manuscript evaluation, it is not clear the protein localization (transmembrane?) since the protein`s end is very close to the virus membrane surface. All experiments demonstrated the protein without being anchored to the membrane, letting the interaction site always be exposed. If the protein is linked to the membrane, how would the site be exposed due to the limited space between it and the virus structure?

Thank you for these insightful comments. As you pointed out, the H3 protein, particularly the helical domain at the C-terminal, is indeed located close to the membrane, which could limit the available space for H3 binding. To investigate this further, we modeled the full-length H3 protein in the context of the membrane and performed molecular dynamics (MD) simulations to assess the available space. Our results show that there is more than 1 nm of space between the helical domain and the membrane, which should be sufficient for potential heparan sulfate (HS) binding (see Figure 1E, and Figure S1D&E in the revised manuscript).

Minor corrections:

Line 31: "is an emerging zoonotic pathogen" should be revised to reflect that Mpox is a re-emerging virus, given its history of causing outbreaks, such as in 2003.

Line 71 and Line 75: Adding an explanation of "Mg binding sites" and "GAG motifs" would enhance reader understanding, as these represent important points in the study. The current positioning of Figure 1 causes some confusion for the reader.

Line 111: High score? What controls were used for the protein? Are there known inhibitors of H3? If so, why weren't they tested for structure comparison? Additionally, what about other molecules that H3 binds to, such as UDP-Glucose, as demonstrated in the base article for the Vaccinia virus H3 protein available in the PDB?

Figure 2B: Improve the legend, as the colors of the lines are not clear.

Thank you for your instructive comments. We have addressed most of them in the revised manuscript.

Regarding the "high score," AlphaFold2 provides a confidence score for its protein structure predictions, with a maximum score of 100. A score above 80 indicates a high level of confidence in the prediction.

There are known inhibitors (such as antibodies) of H3, and while the sequence is available, no structure has been reported so far. Previous s NMR titration measurements have shown that UDP-glucose binds to H3, but no structural data for the complex exist. To date, the only available crystal structure is of a truncated H3, which does not include the helical domain we identified from VAVC.

Reviewer #2 (Recommendations For The Authors):

The text described in the result section does not match the text presented in Figures. So, it is not easy to see what are the authors referring to when they mention the Figure. For example, the text referring to Figure S8 mentions the GB1 domain and the Cohesin module, but these are not mentioned in Figure S8.

I do not understand the results presented in Figure 5B. It is not clear to me, from the Figure legend nor after reading the Material and Methods, how this experiment was done. Specifically, what is plotted on X, is it the amount of inhibitor or the amount of protein? These things have to be checked through the manuscript.

It would be interesting to confirm if the inhibition of infection is based on the inhibition of viral binding to the cells. This should not be complicated to realize, and it could provide evidence for the mechanism of action.

Extensive use of terms like "this domain" is not good in this type of article, like in lines 207, and 211. It is not always clear to what domain are authors referring to, so it may be much better to mention the domain in question by the exact name.

Line 337, If I am not mistaken dilutions are serial not series.

Line 613, in methods. Please use g force instead of rpm, it is more informative. Even if it is just to pellet cells.

Thank you very much for your instructive comments. We have addressed most of them in the revised manuscript. For instance, the immobilization of the GB1 domain and the cohesin module is now mentioned in Figure S9. Additionally, in the previous Figure 5B, the "x" represents the concentration of the inhibitor. Serial and g force is updated.

Reviewer #3 (Recommendations For The Authors):

Line 190

Did you mutate all the amino acids at the same time? What was the impact of all these mutations on the structure of the helical region? Or if you modeled the protein again after replacing these 7 amino acids, did you find that there was no difference? Regardless of your answer, you must include a superposition of the mutated structure and the wt.

Thank you for the insightful comment. We have now also predicted the structure of the serine mutant using AlphaFold2 (AF2). As expected, the helical domain structure remains largely preserved with only minor differences. We have included these results in Figure S6, as suggested.

Figure 2D

In this graph, the authors should indicate the ΔG as a negative value. In fact, the graph does not match the text.

Thanks for the reminder, it is corrected in the graph

Figure 4B

Is the difference in binding force significantly different? 28.8 vs 33.7 pN

The absolute difference in binding force is not large (~5 pN). However, for a system with a relatively low binding force, this difference is significant. Specifically, the 5 pN difference accounts for approximately a 14% reduction in binding force. We have included this percentage in the revised manuscript.

Figure 5

If AI-PoxBlocks723 was the only peptide effective in inhibiting viral infection of MPOX and other related viruses but not with 100% effectiveness, do you think this could be a consequence of a low interaction efficiency or the existence of a different receptor? Or a secondary region of binding in the H3? Can you argue about this?

It has been proposed that there are other adhesion proteins for MPXV, such as D8, in addition to H3. We believe this accounts for the observed less-than-100% effectiveness.

The use of peptides as "inhibitory tools" could have an interesting effect in vitro, however, in vivo the immunological response against the peptide will reduce/eliminate it, how you may optimize the "drug" development with this system, as you state in line 387.

Thank you for your thoughtful comment. You are correct that the use of peptides as inhibitory tools could induce an immune response in vivo, which might limit their effectiveness over time. To optimize this approach for drug development, conjugate the peptides with carrier molecules, such as liposomes, nanoparticles, or dendrimers, which can protect the peptides from immune detection and improve their delivery to target cells. This could allow for more controlled and sustained release of the peptide in vivo, reducing the chances of immune clearance. We have added this discussion in the revised manuscript.

Version 2

Abstract

Abstract

Viral adhesion to host cells is a critical step in infection for many viruses, including monkeypox virus (MPXV). In MPXV, the H3 protein mediates viral adhesion through its interaction with heparan sulfate (HS), yet the structural details of this interaction have remained elusive. Using AI-based structural prediction tools and molecular dynamics (MD) simulations, we identified a novel, positively charged α-helical domain in H3 that is essential for HS binding. This conserved domain, found across orthopoxviruses, was experimentally validated and shown to be critical for viral adhesion, making it an ideal target for antiviral drug development. Targeting this domain, we designed a protein inhibitor, which disrupted the H3-HS interaction, inhibited viral infection in vitro and viral replication in vivo, offering a promising antiviral candidate. Our findings reveal a novel therapeutic target of MPXV, demonstrating the potential of combination of AI-driven methods and MD simulations to accelerate antiviral drug discovery.

Introduction

The mpox virus (MPXV), a zoonotic pathogen within the Orthopoxvirus genus, has emerged as a global health concern. This genus also includes the Variola virus (VARV), the causative agent of smallpox, and Vaccinia virus (VACV), which has been used as a live vaccine against smallpox1, 2 (Fig. 1A). Historically endemic to Central and West Africa, MPXV gained international attention during the global outbreaks in 2022 (clade IIb) and 2024 (clade Ib), leading the World Health Organization (WHO) to declare it a public health emergency of international concern twice within three years3-5. Although MPXV has a lower fatality rate than smallpox (2-10% versus 30%), the absence of specific antiviral treatments continues to pose significant health risks6-11.

Structural Analysis of the H3 Protein in Monkeypox Virus

(A) Phylogenetic tree depicts the evolutionary relationships of H3 within the Poxviridae family, highlighting MPXV (blue circle), VARV (red circle), and VACV (green circle). (B) Schematic of MPXV adhesion to the cell surface. Viral particles bind to cell surface via specific interaction such as between adhesion protein H3 and HS, followed by membrane fusion mediated by the fusion complex, allowing entry into the host cell. (C) The amino acid sequence of MPXV H3 shows the newly discovered helical structure (240-282, highlight in yellow), the Mg(II) (green) binding site and other potential HS binding motifs (blue underlines). (D) AF2 prediction of MPXV H3 structure. The left and right panels show different orientations of the H3 structure (rotated 180°). The blue region corresponds to H3(1-239), which has a homologous crystal structure (VACV H3, PDB code: 5EJ0). The yellow region represents the AF2-predicted structure of H3(240-282), which remains unresolved in the crystal structure of the homologous protein. All potential GAGs-binding motifs are highlighted with a yellow background. (E) MD simulation snapshot of H3 on a DPPC membrane. H3 is anchored to the membrane through its transmembrane region (residues 283-306, in gray).

Viral adhesion to host cells is a critical initial step in the infection process for many viruses. Proteins that mediate these interactions have been identified as essential therapeutic targets, offering promising opportunities for antiviral therapies. In orthopoxviruses such as monkeypox virus (MPXV), the H3 protein plays a pivotal role in facilitating viral adhesion by binding to cell-surface heparan sulfate (HS)12-14, a glycosaminoglycan essential for viral entry. H3-specific neutralizing monoclonal antibodies have shown to protective effects for rabbits and immunodeficient mice against lethal poxvirus infections15. And H3 is a key component in recent mpox vaccines16, 17 Despite the central role of H3 in MPXV infection and protection, the structural details of its specific interaction with HS remain elusive.

Understanding the molecular mechanisms underlying the H3-HS interaction is critical for advancing antiviral drug development. H3 is highly conserved across poxviruses (Fig. 1A), including variola and vaccinia viruses, making it a potential target for broad-spectrum therapies. The recent resurgence of MPXV infections has underscored the urgent need for specific antiviral treatments. Given the conserved nature of H3 across all known MPXV lineages, including Clade Ib, which has been linked to the 2024 outbreak, designing therapies that target H3 could offer a robust solution to current and future outbreaks.

However, efforts to resolve the complete structure of H3-HS complex, have been hampered by the dynamic and flexible nature of HS and the tendency of H3 self-cleavage. Classic structural biology techniques, such as X-ray crystallography, have been unable to capture the full interaction between H3 and HS due to these complexities. Therefore, alternative approaches are required to fully elucidate this critical interaction.

To address these challenges, we employed advanced AI-based protein structural prediction and generation tools, including AlphaFold2(AF2) and RFdiffusion, alongside classic molecular dynamics (MD) simulations. These state-of-the-art computational techniques enabled us to identify key HS-binding sites within H3 and to characterize the binding mechanisms of the previously uncharacterized α-helical domain18-22. Corroborated by various experimental methods23-26, the use of AI-driven tools proved pivotal in overcoming the limitations of traditional structural methods, allowing us to predict and validate the structure-function relationship of the H3-HS interaction. A powerful approach for unraveling complex protein-glycan interactions is demonstrated, offering new pathways for the development of broad-spectrum antiviral therapies27, 28.

Result

Identification of a Novel Helical Domain for HS Binding in H3

The H3 protein of MPXV is a 324-amino-acid transmembrane protein conserved across all 12 identified orthopoxviruses. Previous studies have shown that its extracellular domain (residues 1-282) binds to cell-surface heparan sulfate (HS)29, 30, facilitating viral entry in conjunction with other adhesion proteins and the fusion protein system (Fig. 1B)31. Inhibiting the HS binding to H3, therefore, represents a promising strategy for developing broad-spectrum antiviral therapies against orthopoxviruese. While most structure of the extracellular region of Vaccinia virus H3 (residues 1-240; PDB code: 5EJ0) has been solved, the dynamic and heterogeneous nature of HS as a biopolymer complicates the identification of precise binding sites.

To identify HS binding sites on the H3 protein, we first performed a comprehensive sequence and structural analysis. As a highly negatively charged glycosaminoglycans, it is believed that HS interacts with positive charge residues on proteins by electrostatic interaction. Previous studies have indicated that the positively charged Mg (II) ion in H3 is critical for HS binding. Therefore, we focused on regions rich in basic amino acids—lysine (K), arginine (R), and histidine (H). Given that only the H3 structure from VACV was available28, we used AlphaFold2 (AF2) to model the MPXV H3 structure (Fig. S1A), Our analysis revealed three potential HS binding motifs rich in positively charge residues: Motif 1 (K95, R96, R100), Motif 2 (K141, K146, K147), and Motif 3 (K161, K163)32, all located within the homologous structure of VACV H3 (Fig. 1C, D).

Interestingly, we also identified a cluster of seven additional positively charged residues (R242, H247, R248, K253, R259, K266, R267) in the protein C-terminal region. This region had not been previously structurally characterized due to its self-cleavage during sample preparation for X-ray diffraction studies, but was retained in all other biochemical and viral assays28-30. AF2 predicted these residues to form a three-helical domain with high confidence (pLDDT score: 82.9 for overall structural; 80.6 for the helical domain, 100 for the maximum score), suggesting it may serve as an HS binding site as a structured domain (Fig. 1D). We also tested other models, such as ESMFold (pLDDT:70.6; 50.7) and RoseTTAFold2 (pLDDT: 70.2; 50.7) (Fig. S1B&C) 33, 34. They all predict a helical structure. But their results are different in detail and had lower confidence scores. Thus, we used AF2 for further studies. Another concern is that the helical domain near the virus surface. Thus, we modeled the complete structure of H3 in virus using a model DPPC (dipalmitoylphosphatidylcholine) membrane. It showed a dynamic movement between the helical domain and the membrane, with a maximum distance exceeding 1 nm (Fig. 1E, Fig. S1D&E, Supplementary Video 1). Thus, enough space be present for possible HS binding.

Flexible molecular docking and MD simulations identified the exact HS binding sites in the H3 protein.35, 36. We used a 20-repeat HS unit with a specific composition for detailed analysis ([IdoA2S-GlcNS6S-IdoA-GlcNS(3,6S)]5) (Fig. S2). A 500 ns MD simulation established the equilibrium conformation of H3, and subsequent docking with AutoDock Vina identified four high-scoring conformations for each motif (Fig. 2A, Fig. S3). The docking score calculated by the software reflects the total energy of interactions between molecules. Lower scores (more negative values) indicate more stable binding between the molecules. To address the dynamic nature of HS, we extended our simulations to 3 times additional 1000 ns MD simulations for each conformation. RMSD (Root Mean Square Deviation) analysis demonstrated notably stable binding for HS to both Motif 1 and the newly identified helical domain (Supplementary Video 2), with RMSD values consistently below 4 nm, indicating minimal structural fluctuations during interaction (Fig. 2B).

Identification of HS Binding Motifs in H3 by MD Simulation and Docking

(A-B) Cartoons show docking results of HS with Motifs 1, 2, 3, and the helical domain, respectively. Panels (B) show RMSD values from 1µs MD simulations, color-coded to match the configurations in (A). (C) Schematics illustrates the reaction coordinate in umbrella sampling, highlighting HS dissociation from H3 with a green force curve. (D) Histogram shows the binding free energies for HS-H3 interactions in motifs 1, 2, 3, and the helical domain. (E) A free energy landscape map from a 1000 ns REMD simulation shows HS binding configurations to the helical and Mg(Ⅱ) regions. (F) Panel provides salt bridge formation statistics between HS and H3’s basic amino acids, with a bar chart of average formations. (G) A surface plot shows the frequency of salt bridge formations within H3, with areas of frequent formations in blue. (H) Detailed views of HS-H3 interactions, with the left image showing salt bridges and the right image displaying electrostatic interactions with Mg(Ⅱ). This panel illustrates the impact on HS binding stability to H3 following the removal of Mg(Ⅱ) during the simulation. (J) The effects of mutating all basic amino acids in the helical domain on the binding stability of HS are shown. (K) The “palm-binding” model is depicted where HS is secured by the helical “fingers” and interacts with the Mg(Ⅱ)-bound “palm.”.

Moreover, we quantitatively analyzed the binding free energy between H3 and HS in different conformational states using umbrella sampling (US) techniques. A harmonic potential was applied to HS and gradually pulled away from H3, generating a series of reaction coordinates (Fig. 2C, Fig. S4, Supplementary Video 3). The green force-extension curve represents the force during this stretching process. This curve, along with the corresponding PMF (Potential of Mean Force) variation shown in Figure S4, demonstrated the free energy changes throughout the umbrella sampling process, revealing that the helical domain possesses the highest binding affinity with a ΔG of -45 kCal/mol (Fig. 2D).

Previous studies have suggested the critical role of the Mg(II) ion in stabilizing HS binding, where its removal reduces binding efficiency29. Indeed, the α-helical domain is very close to the Mg(Ⅱ)-bound region in H3. Notably, these two parts appear to form a distinct cavity-like binding pocket for HS (Fig. 1E, Fig. 2E). Thus, we conducted a 1000 ns REMD (Replica Exchange Molecular Dynamics) simulation of the H3-HS complex to further understand the binding dynamics. REMD, an advanced sampling technique, helps overcome the kinetic barriers typically encountered in conformational transitions by utilizing a series of temperature-controlled replicas to enhance the exploration of the conformational landscape37. For this study, we set up 64 replicas over a temperature range from 310K to 387K, cumulatively simulating for 64 µs. This comprehensive simulation captured the HS molecule progressively binding deeper into a cavity formed by the α-helical domain, eventually interacting with the Mg (II) ion as hypothesized (Fig. 2E, Supplementary Video 4).

The free energy landscape, mapped out from these simulations, shows HS transitioning from a stable interaction with the helical domain across an energy barrier into the Mg (II)-enhanced cavity. This dual interaction—first with the helical domain and then with the Mg(II) site—was consistently observed to stabilize the complex further, evidenced by a significantly lower binding free energy of -67 kCal/mol (Fig. S4)38. These results underscore the dynamic nature of the H3-HS interaction and validate our model of sequential binding, which could be critical for designing inhibitors that target these specific interactions.

Given the crucial role of electrostatic interactions in the HS-H3 binding process, we conducted an exhaustive analysis of salt bridge formations across all MD simulation trajectories. This analysis focused on the formation of salt bridges over time between the basic amino acids (R, K, H) of H3 and HS, with brighter areas indicating more frequent formation of salt bridges and robust interaction (Fig. 2F, Fig. S5). A bar graph accompanied these maps, quantifying the average number of salt bridges each amino acid formed during the simulations, further illustrating the intensity and frequency of these interactions. Our findings demonstrated a high affinity of HS for the helical domain of H3, with basic amino acids in this region forming more salt bridges compared to other parts of the protein. The surface hotspot map of H3 (Fig. 2G) visually highlighted these interactions, with varying shades of blue representing the frequency of salt bridge formation, emphasizing their distribution and intensity. In the helical domain, a focus view showed that the 1st, 7th, 9th, and 18th sugar units of HS formed salt bridges with R248, R242, R267, and R259 of H3, respectively (Fig. 2H, left), while the sulfate group on the 2nd sugar side chain of HS bound to Mg(II) (Fig. 2H, right).

To further prove these findings, we performed two control experiments. First, we removed the Mg(Ⅱ) ion during MD simulation. According to the results of umbrella sampling, we extended the simulation of the most stable H3-HS binding conformation to 1000 ns, then removed the Mg(Ⅱ) ion and continued the simulations. The resulting increase in RMSD fluctuations confirmed the stabilizing role of Mg(Ⅱ) in the binding process (Fig. 2I, Supplementary Video 5). Second, we mutated all seven positively charged amino acids in the helical domain to serine, and name it H3 (uncharged). While AF2 predicted that the helical structure would remain intact (Fig. S6), a significant alteration in HS binding dynamics was observed, supporting the critical role of these charged residues in stabilizing the interaction (Fig. 2J, Supplementary Video 6).

Our interaction model revealed that the α-helical domain of H3 functions like a thumb, guiding HS into a stable binding position, while the Mg(Ⅱ) region acts like a palm, securing HS in place. This dynamic interaction allows HS to transition from an initial surface-level binding to being deeply anchored within the protein cavity, illustrating a complex yet well-organized binding process (Fig. 2K).

Bioinformatic analysis of the H3 protein across the Poxviridae family highlights its evolutionary conservation and its significance in HS binding. Given the crucial role of electrostatic interactions in the H3-HS binding process, we analyzed the charge distribution within the helical domain. To date, H3 proteins have been discovered in 66 of 118 known Poxviridae species. Using NCBI’s BLAST, we retrieved sequences of these 66 H3 proteins and performed multiple sequence aligment39. A heatmap detailing the side-chain charge distribution at a physiological pH of 6.5—mimicking the microenvironment of H3-HS interaction—revealed a significant concentration of positive charges at the helical domain (Fig. 3A). Furthermore, structural predictions using AF2 showed that all these 66 sequences generally share a similar architecture to MPXV H3, particularly in the presence of the α-helical domain (Fig. S7).

Charge Characteristics and Structural Analysis of H3 Protein

(A) The heatmap shows the amino acid charge distribution in the H3 protein across 66 Poxviridae viruses, following multiple sequence alignment. Blue indicates areas with more positive charges, and red indicates more negative charges. The accompanying curve shows the average charge of all amino acids in Poxviridae H3. (B) Logo plot of the amino acid sequence of the helical region, highlighting the conservation of basic amino acids at specific positions. (C) The surface charge analysis of H3 from the MPXV, VACV, VARV, and cowpox viruses (CPXV) with the helical domain showing a significantly positive charge. (D) Schematic of the single-molecule force spectroscopy unfolding experiment on H3, illustrating the unfolding process of the helical domain (yellow) followed by the main body (blue). (E) Representative curves of H3 unfolding, color-coded to show the helical region (yellow), main body (blue), and full-length (purple) unfolding. (F) Histograms depict the force spectroscopy signals for helical domain, main body, and full-length unfolding, with ΔLc statistics provided. The inset shows a Gaussian fit of unfolding forces.

Notably, this domain consistently exhibited an accumulation of basic amino acids such as lysine, arginine, and histidine, which are essential for binding the negatively charged HS. These amino acids, including residues identified as 358, 361, 367, 373, 379, 386, and 387 in Poxviridae H3 (global position in alignment), corresponding to residues H239, R242, R248, K253, R259, K266, R267 in MPXV H3, show high conservation (Fig. 3B, Fig. S8). Surface charge analysis of H3 proteins from four orthopoxviruses, including MPXV, VACV, VARV, and CPXV confirmed the predominance of strong positive charges in their α-helical domains (Fig. 3C), crucial for effective binding to HS. This consistent feature across different viruses underscores the evolutionary importance of the helical domain, suggesting a universal mechanism in orthopoxviral adhesion that could be targeted in antiviral strategy.

Experimental Confirmation of the Structure and Function of α-Helical Domain

To validate the computational predictions, we first conducted experiments using atomic force microscopy-based single-molecule force spectroscopy (AFM-SMFS). This technique, widely used for studying protein (un)folding, allowed us to directly assess the structure and stability of the α-helical domain within the H3 protein40-42. We engineered the H3 construct with two GB1 domains, each 18 nm in length upon unfolding, and one Cohesin module for precise measurement (Fig. 3D)43-45. These proteins are immobilized on the surface by click chemistry and enzymatic ligations (Fig. S9)46-48. During the experiment, the coated AFM tip with GB1 and Dockerin initiated a Coh-Doc interaction, producing a characteristic sawtooth-like force-extension curve as it was stretched (Fig. 3D&E). Notably, two peaks corresponding to the unfolding of the α-helical domain were observed. The first peak showed a contour length increment (ΔLc) of 13 nm, closely matching the theoretical unfolding length for the 42 amino acid-long α-helical domain of H3 (residue 240-282, 42aa*0.36 nm/aa-2.6 nm)49. The measured unfolding force was 67.4 ± 2.1 pN (mean ± SEM, n=139), while the second peak indicated the unfolding of the remaining structure of H3, with a ΔLc of 84 nm (Fig. 3F). The total unfolding, represented by a cumulative ΔLc of 97 nm, confirmed the structured and stable nature of the α-helical domain within H3.

We further explored the functional role of the α-helical domain in mediating H3’s binding to HS using AFM on live cells50, 51. The H3 protein was linked to the AFM tip and approached cultured CHO-K1 cells, which express HS, mounted on a Petri dish. Inverted microscopy was used to precisely control the positioning of the AFM tip. Upon contact, the interaction between H3 and HS was initiated, and the binding event was monitored by retracting the tip to generate a force-extension curve. A distinct peak, corresponding to H3-HS interaction, was observed with a dissociation force of 33.7 ± 0.2 pN and a binding probability of 43% (Fig. 4A, B). Repeated trials across different cell surface areas produced a detailed force map, quantifying the distribution of HS unbinding forces and further confirming the critical role of the helical domain in viral adhesion (Fig. 4C, D).

Analysis of the Helical Domain’s Interaction with HS at Cellular Level

(A) Schematic of the cell force spectroscopy experiment setup shows three scenarios: wild-type H3 on an AFM tip interacting with HS on CHO-K1 cells, mutation of all basic amino acids in the H3 helical region to serine, and cells treated with HS hydrolase to remove surface HS before testing. (B) Force-extension curves depict interactions for the wild type, mutant, and control groups, marked with blue and red asterisks for dissociation events. An inset shows the optical microscope positioning the AFM probe. (C) Histograms of dissociation signals comparing the wild type, mutant, and control groups, with an inset detailing the surface distribution of dissociation forces. (D) Statistical graph showing binding probabilities for different groups, highlighting significant differences determined by t-tests (p<1e-5). (E) FCM results illustrate interactions of wild-type (WT) and Uncharged H3 fused with eGFP with CHO-K1 cells, alongside GFP control (green) and cell-only control (Blank, grey). (F) Statistical analysis of FCM data, showing significant differences between groups as determined by t-tests. *****, P<1e-5

Moreover, we performed mutagenesis studies on the α-helical domain by replacing all positively charged residues with serine. AFM experiments on this variant H3(uncharged) showed a detectable force peak with a lower unbinding force of 28.8 ± 0.2 pN, compared to 33.7 ± 0.3 pN for the wild type—representing a 14% reduction (ΔForce > 2*SEM, 95% CI)—and a significant decreased HS binding probability from 43% to 25% (Fig. 4C, D). These results suggest that the basic amino acids in the helical domain are crucial for HS binding. To confirm the specificity of the interaction, CHO-K1 cells were treated with heparinase II, an enzyme that degrades HS. This treatment further reduced the dissociation force to 23.0 pN and the binding probability to 12.6%, confirming that the observed interactions were specific to the HS-H3 complex.

Further validation was carried out using flow cytometry (FCM). We fused eGFP, a fluorescent reporter protein, to the C-terminus of both the H3(WT) and the H3(uncharged). Following incubation of these fusion proteins with cells, FCM analysis showed a significant reduction in mean fluorescence intensity in cells treated with H3(uncharged) compared to H3(WT) (Fig. 4E). Statistical analysis of multiple experiments highlighted a significant difference between the two (Fig. 4F). These findings, along with the results from MD simulations and AFM force spectroscopy, underscore the importance of the positively charged amino acids in the α-helical domain of H3 for efficient HS binding.

The Helical Domain Serves as a Target for H3 Infection Inhibition

Targeting the α-helical domain of H3 as H3 binding sites, we de novo designed a series of inhibitors. Using deep-learning model RFdiffusion, we designed a series of inhibitors for the H3-HS interaction and named it AI-PoxBlock (Fig. 5A). Sequence recovery for each series was performed on 1000 backbones using ProteinMPNN, generating ten sequences per backbone. These designs were further validated through complex structure predictions using AF2-multimer, and candidates were selected based on PAE (Predicted Aligned Error) values below 7 and iPTM (interface predicted Template Modelling) scores above 0.9. The predicted structures of these candidates also exhibited RMSD values of less than 2 nm when compared to the RFdiffusion-generated structures (Fig. S10). To assess the binding capabilities of these inhibitors, we conducted 500 ns MD simulations, analyzing RMSD trajectories. Five inhibitors—AI-PoxBlock302, 602, 614, 723, and 761—demonstrated stable RMSD values under 0.4 nm, confirming their potential for stable binding to the helical domain of H3 (Fig. S11, S12, Supplementary Video 7).

Inhibitor Design Targeting the H3-HS Interaction and Efficacy of AI-PoxBlock723

(A) Diagrams depict protein inhibitors designed to target the H3 helical region, created using RFdiffusion. Sequences capable of folding into the target scaffold structures were generated using ProteinMPNN, and were validated through AF2, followed by 500 ns MD simulations for structural stability and interaction scoring. (B) FCM analysis demonstrates the inhibitory effect of AI-Poxblock723 at various concentrations (0, 10, 50 and 100 μM, x axis). The control group consisted of 2 μM H3-eGFP without the inhibitor, with the relative fluorescence intensity normalized to 1. (C) BLI confirms direct interaction between AI-Poxblock723 and the H3 helical domain. (D) Graphs display the inhibitory effect of the indicated AI-Poxblocks on MPXV infection of Vero E6 cells, with quantitative analysis of virus-infected foci provided on the right. (E-F) BALB/c mice were infected intranasally with 4 × 105 FFU MPXV and treated with single dose of AI-PoxBlock723 (10 mg/kg) or PBS immediately after challenge (n = 5). Infectious MPXV particles and MPXV viral loads in mice lungs at 4 dpi were determined by focus-forming assay (E) and qPCR (F).

Although AI-PoxBlock302 could not be successfully expressed, the remaining four inhibitors underwent further testing in FCM to evaluate their efficacy in inhibiting H3 binding to CHO-K1 cell surfaces. AI-PoxBlock723 stood out, significantly reducing H3 binding at a concentration of 10 μM, while the other inhibitors displayed no tendency to inhibit H3 binding to HS on cell surfaces (Fig. 5B, Fig. S13). Further analysis using biolayer interferometry (BLI) confirmed that AI-PoxBlock723 binds to H3 with an equilibrium dissociation constant (KD) of 9 μM (Fig. 5C). To ensure specificity, we also assessed the interaction between AI-PoxBlock723 and a truncated version of H3 (residues 1-239), lacking the α-helical domain. BLI results indicated a marked reduction in binding, reinforcing the specificity of AI-PoxBlock723 for the helical domain (Fig. S14). Circular Dichroism spectroscopy also confirmed that AI-PoxBlock723 predominantly consists of α-helices, consistent with its design specifications (Fig. S15).

To evaluate the antiviral activity of AI-PoxBlocks, the inhibitors were serially diluted and incubated with the clade II MPXV isolate SZTH42 before infecting Vero E6 cells. After overnight culture, virus infection foci were determined as previously described.52 Notably, AI-PoxBlock723 showed an antiviral effect with the half maximal inhibitory concentration (IC50) of 88.6 μM, whereas other AI-PoxBlocks exhibited no inhibitory efficiency (Fig. 5D). Given the high conservation of the H3 helical domain across the four orthopoxviruses pathogenic to humans (Fig. S16), we also assessed AI-PoxBlock activity against VACV infection. Consistently, AI-PoxBlock723 is also effective for inhibiting VACA infection in vitro with a similar IC50 of 86.7 μM (Fig. S17), suggesting the helical domain as a universal antiviral target for orthopoxvirus.

To evaluate the efficacy of AI-PoxBlock723 in vivo, groups of BALB/c mice were challenged with 4 × 105 FFU (focus forming unit) MPXV/SZTH42 as we previously reported53. Mice were administered intraperitoneally with single dose of AI-PoxBlock723 (10 mg/kg) or PBS immediately after MPXV challenge, and were sacrificed 4 days post-infection (dpi) for the determination of pulmonary MPXV titers. The MPXV infectious particles and genome copies in lungs of the AI-PoxBlock723 treated mice were significantly lower than that in the control mice (Fig. 5E-F, Fig. S18).

These results demonstrate the efficacy of the designed inhibitors targeting the H3 helical domain and validate the crucial role of this domain in viral adhesion, providing a promising foundation for the development of novel antiviral agents.

Discussion

Our study provides new mechanistic insights into MPXV infection by identifying the α-helical domain of the H3 protein as a critical mediator of heparan sulfate (HS) binding. This domain, previously uncharacterized, plays a pivotal role in viral adhesion to host cells, a key step in MPXV pathogenesis. Using a combination of AI-based tools and molecular dynamics (MD) simulations, we successfully identified the specific interaction sites between H3 and HS, offering a novel target for antiviral drug design54.

Previous studies have highlighted the importance of viral adhesion proteins in orthopoxvirus infection, but the specific mechanisms underlying the H3-HS interaction remained unresolved. Our findings provide the first detailed structural insight into this interaction, confirming the importance of the electrostatic nature of H3-HS binding. These results align with prior reports indicating the role of HS in viral entry, but advance the field by identifying specific residues within the α-helical domain that are crucial for binding.

One of the strengths of this study lies in the application of advanced AI-based tools, such as AlphaFold2 and RFdiffusion, alongside MD simulations. These cutting-edge computational approaches allowed us to overcome the limitations imposed by the dynamic and flexible nature of HS. In addition, HS serves as a cellular ligand for various viruses55-57, underscoring the importance of studying its interactions with viral adhesion proteins. By predicting the structure-function relationship of the H3-HS interaction, we were able to pinpoint key binding residues and mechanisms that were previously inaccessible. This demonstrates the growing potential of AI in structural biology, particularly for the complex protein-glycan interactions56, 58,59.

The identification of the α-helical domain in H3 as a key player in HS binding offers a promising new target for antiviral drug development17, 60-67. Given the conservation of this domain across orthopoxviruses, including variola and vaccinia viruses, targeting this region could lead to the development of broad-spectrum antivirals effective against multiple orthopoxviruses68. Our study highlights the power of AI-driven drug design, as demonstrated by the promising results of AI-PoxBlock723, which showed efficacy in inhibiting viral infection. Further optimization and testing of these inhibitors in clinical models will be critical steps toward developing effective antiviral therapies. For example, conjugate the peptides with carrier molecules, such as liposomes, nanoparticles, or dendrimers, which can protect the peptides from immune detection and improve their delivery to target cells.

Looking ahead, the methodologies and findings from this study are likely to catalyze further investigations into viral adhesion mechanisms, not only in orthopoxviruses but also across other viral families that utilize similar adhesion strategies. Although challenges remain, we hope our discovery will attract structural biologists to resolve this domain’s structure in H3, and even the structure of H3-HS complex, to further enhance drug design efficiency for this target. Although this domain itself cannot express, we chemically synthesized it (residues 240-279). CD measurement on this fragment indeed showed an α-helical structure (Fig. S19). As AIdriven approaches continue to evolve, they are poised to play a transformative role in advancing our understanding of viral pathogenesis and revolutionizing the future of drug discovery.